---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Natalie Stevens"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
health_inspection_scores <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- health_inspection_scores |> mutate(compliance_mean = mean(compliance_score) , compliance_sd = sd(compliance_score) , compliance_min = min(compliance_score) , compliance_max = max(compliance_score))
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
compliance_stats |>
  ggplot(aes(x = compliance_score)) +
  geom_histogram(binwidth= 5, fill = "lightblue", color = "black") +
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype= "dashed", linewidth = 1) + geom_vline(aes(xintercept = mean(compliance_score + sd(compliance_score)), color = "orange", linetype = "dashed")) +
  geom_vline(aes(xintercept = mean(compliance_score) - sd(compliance_score)), color = "orange", linetype = "dashed") +
  labs(
    title= "Health Inspection Compliance Scores in Montgomery County",
    x = "Compliance Scores",
    y = "Number of Establishments"
  )
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**
  Most establishments have a health inspection compliance score above 90 and within one standard deviation of the mean. Notably many fall at exactly 90 or 100 with fewer between. Newsworthy scores may be the lowest of the compliance scores that fall well below the mean, which could be used to create a story about food safety hazards. A next step could be looking further into businesses with low scores and speaking to health inspectors about the matter.

---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
WMATA_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
 wmata_stats <- WMATA_daily |> summarize(bus_mean = mean(bus), bus_sd = sd(bus), rail_mean = mean(rail), rail_sd = sd(rail))

```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- WMATA_daily |> mutate(
  mean_rail = mean(rail),
  mean_bus = mean(bus),
  sd_rail = sd(rail),
  sd_bus = sd(bus),
  rail_zscore = (rail - mean_rail) / sd_rail,
  bus_zscore = (bus - mean_bus) / sd_bus
)
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
wmata_zscores |>
  filter(abs(rail_zscore) > 2) %>%
  arrange(desc(abs(rail_zscore)))
```
There are 6 unusual days.

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(42)

sample_ridership <- WMATA_daily |> sample_n(91)

sample_ridership |> summarize(
  bus_sample_mean = mean(bus) ,
  bus_sample_sd = sd(bus) ,
  rail_sample_mean = mean(rail) ,
  rail_sample_sd = sd(rail)
)
```
  The bus means are similar while the standard deviation is smaller in the population, the sample rail mean is higher while the standard deviation is around 8k riders away from the population's. Random sampling is able to yield results that can resemble the actual population's, but larger samples will generally be closer to it. 

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

Unusual days appear to happen most often in January. These are low-ridership days and could indicate low travel right after the new year or loss of ridership from commuters like students, interns, and government workers on break/vacation. The vacation theory is also supported by the December date, which is Christmas Day. Looking into possible issues with the rails during January may also uncover a story topic.

---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
car_thefts_2023 <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- car_thefts_2023 |> mutate(theft_rate = thefts_2023/population * 10000 )
```
A per-capita rate allows thefts to be seen directly in comparison with a county's population. While higher-population counties will generally have more total car thefts, a small county with a high car theft rate is unusual and may be worth writing a story about.

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
md_car_thefts |> summarize(
  median_rate = median(theft_rate) ,
  statewide_thefts = sum(thefts_2023)
)
high_rates <- md_car_thefts %>%
  filter(abs(theft_rate) > 9.580638) %>%
  arrange(desc(abs(theft_rate)))
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)
Twelve counties have car theft rates above the median, including two low-population counties (St. Mary's County and Dorchester County). The three highest rates are for Baltimore City, Prince George's County, and St. Mary's County. The number of thefts has increased in all 12 counties from 2022.

The number of car thefts in twelve Maryland counties has increased from 2022. Baltimore City's and Prince George's County's theft rates have gone up to 123 and 89 per 10000 people respectively, according to a set of crime data spanning across 2022 to 2023.

---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)
I would prioritize comparing the distribution of visitors per day per pool to the mean and figuring out the z-scores for each. I would also want to know the median number of visitors per pool, as higher-performing pools can skew the mean and hide underperforming pools. 

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)
Raw attendance can be attributed to factors like location and surrounding population, which all create an expected level of performance. Less populated areas or areas where more people have home pools would be expected to perform differently than higher population areas with less home pools. Calculating z-scores for each pool in comparison with the population of the area it is in and considering other variables would give an actual look into whether their performances are below expectation.

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)
I would ask if all pools had their samples taken on the same days, which days of the year (and week) they were taken, and how many days per year the pools are open. If samples were taken on different days for different pools, factors like weather, day of the week, time of year, temperature and holidays could impact certain samples. If the sample of days is large enough compared to the total number of days the pools are open, the data should be somewhat reliable, if difficult to use because of a lack of population data to compare it to. 

---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
