---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Natalie Stevens"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

There are no schools in Montgomery County, MD that have a FARMS % of 100, unlike the PA data. There are no noticeable jumps in percentage for MD and schools have a minimum of exactly 4% of student covered by the program. In general this data looks more reliable than PA's based on FARMS distribution alone.


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

The MD data has a stronger relationship between FARMS and test scores than PA, which had no correlation in a slightly positive direction while MD's is negative. Things like data quality, state testing and curriculum standards and population (the PA county has about 300,000 less people than the MD county) could explain the differences between the two counties. 


### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

The combination of the p-value, R-squared, and FARMS pct coefficient show that there is a definite, medium-strength negative relationship between poverty and math scores in the MD county. This model can be used, but the PA model cannot because of the extremely weak relationship it shows.


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

The school with the largest positive residual is Fox Chapel Elementary.
  Less than 10% of the residents at this zip code live under the poverty line and the median income is $119,731. The majority of residents are under 64 and most end up graduating high school.
  The income of the area may help with improving test scores, as children would have more access to things like tutors and schools would be able to afford better teaching materials.
  My first question would ask if the school has any special programs or teaching methods in place that have improved performance in students.

The school with the largest negative residual is Germantown Elementary.
  Less than 9% of the residents at this zip code live under the poverty line and the median income is $115,022. The majority of residents are under 64 and most end up graduating high school.
  Unlike the other school, the residents at this zip code have greater variation in median income, though fewer children live under the poverty line. Nothing in the profile points to a concrete reason as to why this school would be performing worse than expected compared to the other.
  My first question would ask if the school plans to implement anything to improve student performance.
  
  Both zip codes have very similar demographics.

### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

  The amount of FARMS-eligible students in Pennsylvania schools has no relationship to student performance, but does in Maryland. The relationship is real but small, as schools over- and under-perform frequently. I would ask the school board member how they came to their conclusion based on Pennsylvania's statistics. The most interesting school to visit first would be Germantown Elementary, as there are no clear explanations for its underperforming students when a school at a similar zip code is greatly overperforming.
