---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

  Over half of the salaries are over $100k. 8 salary values are 0, which may cause issues for analysis because they are outliers. There are occasionally significant disparities between individual salaries for people with the same job titles, though some are completely uniform.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
#couldn't get summarize() to work properly :(
#these show up in the values tab but not the dataframe
mean_salaries <- mean(wh_salaries$SALARY)
``` 
```{r}
median_salaries <- median(wh_salaries$SALARY)
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
min_salary <- min(wh_salaries$SALARY)
max_salary <- max(wh_salaries$SALARY)
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean and median salaries are similar, off by under `$8000`. The majority of salaries are over `$100k`, with the highest being `$225700` and the lowest being `$0` (which appear to be outliers, seeing as the next-highest salary is `$59070`).

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I would look further into the salary differences I mentioned in task 3 and narrow them down to a few specific job titles to make a more concise story. I would have to look into factors for each individual such as time spent under each title to form a story.
